---
title: "R Package (self-learning): `data.table` - Reshape"
date: "2023-11-15"
output: github_document
---

Tutorial website: https://rdatatable.gitlab.io/data.table/articles/datatable-reshape.html 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-package, message = F}
library(data.table)
```

## Introduction 

The `data.table` package developes (?) two functions to reshape the data tables:   

- `melt()`: *wide to long*;   
- `dcast()`: *long to wide*.    

The implementations are specifically designed with large in-memory data (e.g. 10Gb) in mind. 


## 1. Default functionality

### a) `melt`ing `data.table`s (wide to long)
 
```{r}
s1 <- "family_id age_mother dob_child1 dob_child2 dob_child3
1         30 1998-11-26 2000-01-29         NA
2         27 1996-06-22         NA         NA
3         26 2002-07-11 2004-04-05 2007-09-02
4         32 2004-10-10 2009-08-27 2012-07-21
5         29 2000-12-05 2005-02-28         NA"
DT <- fread(s1)
DT

str(DT)
```


#### - Convert `DT` to *long* form where each `dob` is a separate observation. 

We can specify the following two arguments in the `melt()` function: 

- `id.vars`: specify the set of columns we would like to keep them identical. 
  
- `measure.vars`: specify the set of columns we would like to collapse (or combine) together;   
  - We can also specify *column indices* instead of *names*.

- By default, when one of `id.vars` or `measure.vars` is missing, 
  the rest of the columns are automatically assigned to the missing argument. 
  Shown in the [next example](#1a2). 

For example: 
```{r}
DT.m1 <- melt(DT, id.vars = c("family_id", "age_mother"),
                  measure.vars = c("dob_child1", "dob_child2", "dob_child3"))

DT.m1

str(DT.m1)
```

- By default, the molten columns are automatically named `variable` and `value`.

- By default, `variable` column is of type `factor`. 
  - Set argument `variable.factor` to `FALSE` if you'd like to return a `character` vector instead. 

- `melt` preserves column attributes in result. 


#### - Name the `variable` and `value` to `child` and `dob` respectively. {#1a2}
```{r}
DT.m1 <- melt(DT, measure.vars = c("dob_child1", "dob_child2", "dob_child3"), 
                variable.name = "child", value.name = "dob")
DT.m1 
```

- By default, when one of `id.vars` or `measure.vars` is missing, 
  the rest of the columns are automatically assigned to the missing argument.  

- When neither `id.vars` nor `measure.vars` are specified, 
  all *non-*`numeric`, `integer`, `logical` columns will be assigned to `id.vars`. 



### b) `dcast`ing `data.table`s (long to wide)

#### - Get back to the original data table `DT` from `DT.m1` 

We use `dcast` function to accomplish it.  

```{r}
dcast(DT.m1, family_id + age_mother ~ child, value.var = "dob")
```

- `dcast` uses *formula* interface: `id variables ~ measure variables`   
  - the variables on the *LHS* of formula represents the *id* vars;  
  - those on the *RHS* represents the *measure* vars.

- `value.var` denotes the column to be filled in while casting to wide format.

- `dcast()` also tries to preserve attributes in result wherever posible. 

#### - Starting from `DT.m1`, how can we get the number of children in each family? 

Pass a function to aggregate by in `dcast()` with the argument `fun.aggregate`. 
This is particular essential when the formula provided does not identify single observation for each cell. 

```{r}
dcast(DT.m1, family_id ~., 
      fun.agg = function(x) sum(!is.na(x)), 
      value.var = "dob")
```


## 2. Limitations in current `melt/dcast` approaches 

Here is another example: 
```{r}
s2 <- "family_id age_mother dob_child1 dob_child2 dob_child3 gender_child1 gender_child2 gender_child3
1         30 1998-11-26 2000-01-29         NA             1             2            NA
2         27 1996-06-22         NA         NA             2            NA            NA
3         26 2002-07-11 2004-04-05 2007-09-02             2             2             1
4         32 2004-10-10 2009-08-27 2012-07-21             1             1             1
5         29 2000-12-05 2005-02-28         NA             2             1            NA"
DT <- fread(s2)
DT
```


Suppose we'd likd to combine (`melt`) all the `dob` columns together, and `gender` columns together. 
We may do something like this: 
```{r}
# 1. & 2.
DT.m1 = melt(DT, id.var = c("family_id", "age_mother"))  
# 3. 
DT.m1[, c("variable", "child") := tstrsplit(variable, "_", fixed = TRUE)]
# 4. 
DT.c1 = dcast(DT.m1, family_id + age_mother + child ~ variable, value.var = "value")

DT.c1
str(DT.c1)
```

**Issues:**  

1. We combine both `dob` and `gender` columns together, and then split them again, 
   which is quite "roundabout" and inefficient. 

2. The columns to `melt` may be of different types, as in this case (`character` and `integer`). 
   By `melt`ing them all together, the columns will be coerced in result: 
   in this case, `gender` has been converted to `character` type. 

3. We are generating an additional column by splitting the `variable` column into two columns, 
   whose purpose is quite cryptic. 
   We do it because we need it for *casting* in the next step. 

4. Finally, we cast the data set. But it requires computing the order of the variables in formula, 
   which is much more computationally involved operation than `melt`. 

Well, try `stats::reshape()`!


## 3. Enhanced (new) functionality

### a) Enhanced `melt()`

#### - `melt` multiple columns simultaneously  

We may pass a list of columns to `measure.vars`, 
where each element of the list contains the columns that should be combined together. 
```{r}
colA = paste0("dob_child", 1:3)
colB = paste0("gender_child", 1:3)
DT.m2 = melt(DT, measure = list(colA, colB), 
             value.name = c("dob", "gender"))
DT.m2
str(DT.m2)
```

- We can remove the `variable` column if necessary. 

#### - Using `patterns()` 

The columns we'd like to melt may be distinguished by a common pattern. 
We can use the function `patterns()` to provide regular expressions for the columns to be combined together. 
```{r}
DT.m2 = melt(DT, measure = patterns("^dob", "^gender"), 
             value.name = c("dob", "gender"))
print(DT.m2, class = TRUE)
```

#### - Using `measure()` to specify `measure.vars` via separator or pattern 

If the input columns to melt have regular names, we can use `measure`, 
which allows specifying the columns to melt via a separator or a regex. 

For example:
```{r}
 (two.iris <- data.table(datasets::iris)[c(1, 150)])
```

The iris data has four numeric columns with a regular structure: 
"flower part + . + measurement dimension". 
To specify that we want to melt those four columns, 
we can use `measure` with `sep = "."`, 
which means to use `strsplit` on all column names; 
the columns which result in the maximum number of groups after splitting will be 
used as `measure.var`: 

```{r}
melt(two.iris, measure.vars = measure(part, dim, sep = "."))
```

- The first two arguments to `measure()` in the code above (`part` and `dim`) are used to 
  name the output columns;   
  - the number of arguments must equal the max number of groups after splitting with `sep`. 

If we want two value columns, one for each part, we can use the special `value.name` keyword, 
which means to **output a value column for each unique name found in that group**: 
```{r}
melt(two.iris, measure.vars = measure(value.name, dim, sep = "."))
```

Similarly, 
```{r}
melt(two.iris, measure.vars = measure(part, value.name, sep = "."))
```


Let's go back to the example of the data with families and children.
```{r}
DT
```

We would like to collapse all `dob` columns together, and `gender` columns together. 
Besides, we would like to convert the `child` string values into integers: 

```{r}
DT.m3 = melt(DT, measure = measure(value.name, child = as.integer, sep = "_child"))
print(DT.m3)
```

- Here, columns containg the strings `_child` are melted:  
  `dob_child1`, `dob_child2`, `dob_child3`, 
  `gender_child1`, `gender_child2`, and `gender_child3`. 

- Specified by `value.name`, strings before `_child` are used to name the new columns; 
  we have two new columns named `dob` and `gender`. 

- `child=as.integer` means the second group will result in an output column 
  named `child` with values defined by plugging the character strings from 
  that group into `as.integer`. 


Finally consider the following example 
where we need to define the groups using a regular expression rather than a separator. 
```{r}
(who <- data.table(id = 1, new_sp_m5564 = 2, newrel_f65 = 3))

melt(who, measure.vars = measure(
     diagnosis, gender, ages, pattern = "new_?(.*)_(.)(.*)"))
```

> To learn more about regular expression, here is a 
[quick reference](https://learn.microsoft.com/en-us/dotnet/standard/base-types/regular-expression-language-quick-reference)

When using the `pattern` argument, it must be a Perl-compatible regular expression 
containing the same number of capture groups (parenthesized sub-expressions) 
as the number other arguments (group numbers).  

A more complex example: 
```{r}
print(
    melt(who, measure.vars = measure(
        diagnosis, gender, ages, 
        ymin = as.numeric, 
        ymax = function(y) ifelse(y == "", Inf, as.numeric(y)), 
        pattern = "new_?(.*)_(.)(([0-9]{2})([0-9]{0,2}))"
    )),
    class = TRUE
)
```


### b) Enhanced `dcast` 

Given `DT.m2` as shown above, 
how can we get back to the same format as the original data? 

```{r}
DT.m2
```

#### - Casting multiple `value.var`s simultaneously

We can provide multiple `value.var` columns to `dcast` for `data.table`s 
directly. 

```{r}
## new 'cast' functionality - multiple value.vars 
DT.c2 = dcast(DT.m2, family_id + age_mother ~ variable, 
                     value.var = c("dob", "gender"))
DT.c2
```


#### Multiple functions to `fun.aggregate`. 

Check examples in `?dcast` that illustrates this functionality. 