---
title: "R Package (self-learning): `data.table` - Introduction"
date: "2023-11-14"
output: github_document
---

Tutorial website: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html  

(I copied the contents in the vignette to teach myself, 
thus the following contents are basically the same as those in the vignette.) 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-package}
library(data.table)
```

Installing R package `data.table` in MacOS may encounter a problem of not detecting OpenMP support. 
It indicates that we are using single-threaded mode instead of the faster multi-threaded mode. 
To deal with it, read [this](https://mac.r-project.org/openmp/) and [this](https://github.com/Rdatatable/data.table/wiki/Installation).



## Data {#data}

- Use file reader `fread` to load data directly. 

Following the tutorial, we use [NYC-flights14](https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv)
obtained by [flights](https://github.com/arunsrinivasan/flights) package.

```{r load-data}
input <- if(file.exists("flights14.csv")) {
    "flights14.csv"
} else {
    "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv"
}
flights <- fread(input)
flights
```

Aside: `fread` accepts `http` and `https` URLs directly as well as operating system commands such as `sed` and `awk` output. 
See `?fread` for examples.

## 1. Basics

`data.table` is an R package providing an enhanced version of `data.frame`s. 

### a) Creat a `data.table`

- `data.table()`: create a `data.table` from scratch; similar to `data.frame()` function. 

```{r}
DT = data.table(
    ID = c("b", "b", "b", "a", "a", "c"),
    a = 1:6, 
    b = 7:12, 
    c = 13:18
)
DT
class(DT$ID)
```
  
- `fread()`: create a `data.table` from existing files, as shown in the [Data](#data) section above.

- `setDT()`: convert existing objects such as `data.frame`s and `list`s to a `data.table`. 

- `as.data.table()`: convert existing objects of other structures to a `data.table`.

> Note that: 
>  
> - Unlike `data.frame`s, columns of `character` type are *never* converted to `factors` by default.  
> - Row numbers are printed with a `:`.  
> - `data.table` does not set or use *row* names.  
> - `data.table` automatically prints only the top 5 and bottom 5 rows if the number of rows exceeds the global option `datatable.print.nrows` (default = 100). 


### b) General form of `data.table` syntax 

```
DT[i, j, by]  

##   R:                 i                 j        by
## SQL:  where | order by   select | update  group by
```
**The way to read it is:**   
Take `DT`, subset/reorder rows using `i`, then calculate `j`, grouped by `by`. 

> We might also refer to writing thins inside `DT[...]` as querying `DT`. 

### c) Subset rows in `i` 

#### - Get all flights with "JFK" as the origin airport in the month of June 

```{r}
ans <- flights[origin == "JFK" & month == 6L]
head(ans)
```

- Within the frame of a `data.table`, **columns can be referred to as if they are variables**.  
  We do not need to add prefix `flights$` each time.    
- The *row indices* that satisfy the condition `origin == "JFK" & month == 6L` are computed, 
  and since there is nothing else left to do, all columns from `flights` at selected rows are simply returned as a `data.table`.   
- A comma after the condition in `i` is not required. But `flights[origin == "JFK" & month == 6L, ]` would work just fine.

#### - Get the first two rows from `flights`
```{r}
ans <- flights[1:2]
ans
```

- In this case, the row indices are provided in `i`. 


#### Sort `flights` first by column `origin` in *ascending* order, and then by `dest` in *descending* order 

We can use `order()` to accomplish this. 

```{r}
ans <- flights[order(origin, -dest)] 
head(ans)
```

**`order()` is internally optimized** 

- we can use `-` on a `character` columns within the frame of a `data.table` to sort in *decreasing* order. 


### d) Select column(s) in `j` {#1d}

#### - Select `arr_delay` column, but return it as a vector

```{r}
ans <- flights[, arr_delay]
head(ans)
```

- Since we want *all the rows*, we simply skip `i`.


#### - Select `arr_delay` column, but return as a `data.table` instead. 

```{r}
ans <- flights[, list(arr_delay)]
head(ans)
```

- We wrap the *variables* (column names) within `list()`, which ensures that a data.table is returned.  
- `data.table` also allows wrapping columns with `.()` instead of `list()`. 
  The former is an alias for the latter. They both mean the same.
```{r}
ans <- flights[, .(arr_delay)]
head(ans)
```

> **Tips:**  
> As long as `j-expression` returns a `list`, each element of the list will be converted to a column 
> in the resulting `data.table`. 


#### - Select both `arr_delay` and `dep_delay` columns
```{r}
ans <- flights[, .(arr_delay, dep_delay)]
head(ans)

## alternative
# ans <- flights[, list(arr_delay, dep_delay)]
```


#### - Select both `arr_delay` and `dep_delay` columns and rename them to `delay_arr` and `delay_dep`

Since `.()` is just an alias for `list()`, 
we can name columns as we would while creating a `list` 
```{r}
ans <- flights[, .(delay_arr = arr_delay, delay_dep = dep_delay)]
head(ans)
```


### e) Compute or *do* in `j` 

#### - How may trips have had total delay < 0? 
```{r}
ans <- flights[, sum( (arr_delay + dep_delay) < 0 )]
ans
```

- `data.table`'s `j` can handle expressions as well, i.e., *computing on columns*. 

### f) Subset in `i` and do in `j` 

#### - Calculate the average arrival and departure delay for all flights with "JFK" as the origin airport in the month of June. 
```{r}
ans <- flights[origin == "JFK" & month == 6L, 
               .(m_arr = mean(arr_delay), m_dep = mean(dep_delay))]
ans
```

- We first subset in `i` to find matching *row indices*. We *do not* subset the entire `data.table` corresponding to those rows yet. 
- In `j`, we only uses two columns. And what we have to do is to compute their `mean()`.
  Therefore we subset just those columns corresponding to the matching rows, and compute their `mean()`. 

> Note: `data.table` can see all three main components of the query (`i`, `j`, and `by`) altogether before evaluation, 
> not each separately. 

#### - How many trips have been made in 2014 from "JFK" airport in the month of June? 

```{r}
ans <- flights[origin == "JFK" & month == 6L, length(dest)]
ans
```
Here, `length()` is just a function to count the number of rows; 
its input argument can be any column names other than `origin` and `month`.  

#### Special symbol `.N`: 

We can accomplish the same task by using `.N` as follows:
```{r}
ans <- flights[origin == "JFK" & month == 6L, .N] 
ans
```

- `.N` is a special built-in variable that holds the number of observations *in the current group*. 
  It is particularly useful when combined with `by` as we will see in the next section.   
- Note that we did not warp `.N` with `list()` or `.()`, so a vector is returned. 


### g) Refer to columns by names in `j` (like in a `data.frame`) {#1g}

If you are writing out the column names explicitely, there is not difference  from a `data.frame` 

#### - Select both `arr_delay` and `dep_delay` columns in the `data.frame` way

```{r}
ans <- flights[, c("arr_delay", "dep_delay")]
head(ans)
```

#### - Select the desired columns in a character vector

There are two options: Using the `..` prefix or the `with` argument. 

**Option 1:** using the `..` prefix
```{r}
select_cols = c("arr_delay", "dep_delay")
flights[, ..select_cols] 
```
 
- Similar to Unix terminal, the `..` prefix refers to "up-one-level" command.


**Option 2:** using `with = FALSE` argument
```{r}
flights[, select_cols, with = FALSE]
```

- The argument is named `with` after the R function `with()` because of simlar functionality. 

```{r}
DF <- data.frame(x = c(1, 1, 1, 2, 2, 3, 3, 3), y = 1:8)

## (1) normal way
DF[DF$x > 1, ]

## (2) using with 
DF[with(DF, x > 1), ]
```
- Using `with()` in (2) allows using `DF`'s column `x` as if it were a variable.    
- Similarly, setting `with = FALSE` disables the ability to refer to columns as if they are variables.  
  - `with = TRUE` is the default in `data.table` because we can do much more by allowing `j` to handle expressions.  

We can also **deselect** columns using `-` or `!`. For example: 
```{r, eval = F}
## not run

# return all columns except arr_delay and dep_delay 
ans <- flights[, !c("arr_delay", "dep_delay")]
# or 
ans <- flights[, -c("arr_delay", "dep_delay")]
```

We can also select by specifying starting and end column names:
```{r, eval = F}
## not run 

# returns year, month and day
ans <- flights[, year:day] 
# returns day, month and year (reverse order)
ans <- flights[, day:year]
# return all columns except year, month and day
ans <- flights[, -(year:day)]
ans <- flights[, !(year:day)]
```



## 2. Aggregations

### a) Grouping using `by` 

#### - How can we get the number of trips corresponding to each origin airport? 

```{r}
ans <- flights[, .(.N), by = .(origin)]
ans

## or equivalently using a character vector in `by`
# ans <- flights[, .(.N), by = "origin"]
```

- Grouping by `origin` obtains the number of rows, `.N`, for each group. 

- `by` also accepts a character vector of column names.  

- When there is only one column or expression to refer to in `j` and `by`, se can drop the `.()` notation. 
```{r}
ans <- flights[, .N, by = origin] 
ans
```

#### - How can we calculate the number of trips for each origin airport for carrier code `"AA"`? 

```{r}
ans <- flights[carrier == "AA", .N, by = origin] 
ans
```

- We first obtain the row indices for the expression `carrier == "AA"` from `i`.    
- Using those *row indices*, we obtain the number of rows while grouped by `origin`. 

#### - How can we get the total number of trips for each `origin, dest` pair for carrier code `"AA"`? 

```{r}
ans <- flights[carrier == "AA", .N, by = .(origin, dest)]
head(ans)

## or equivalently using a character vector in 'by'
# ans <- flights[carrier == "AA", .N, by = c("origin", "dest")]
```


#### - How can we get the average arrival and departure delay for each `origin, dest` pair for each month for carrier code `"AA"`? 

```{r}
ans <- flights[carrier == "AA", 
               .(mean(arr_delay), mean(dep_delay)), 
               by = .(origin, dest, month)]
ans
```

- Since we did not provide column names for the expressions in `j`, 
  they were automatically generated as `V1` and `V2`.   
- Note that the input order of grouping columns is preserved in the result. 

### b) Sorted `by`: `keyby` 

#### - How can we directly order by all the grouping variables? 

```{r}
ans <- flights[carrier == "AA", 
               .(mean(arr_delay), mean(dep_delay)),
               keyby = .(origin, dest, month)]
ans
```

- `keyby` automatically orders the result by the grouping variables in increasing order. 
- **Keys:** `keyby` does more than *just ordering*. Learn more about `keys` in the 
  [*"Keys and fast binary search based subset"*](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html) vignette. 


### c) Chaining
Let's reconsider the task of getting the total number of trips for each `origin, dest` pair for carrier code `"AA"`: 

```{r}
ans <- flights[carrier == "AA", .N, by = .(origin, dest)]
```


#### - How can we order `ans` using the columns `origin` in ascending order, and `dest` in desceding order? 
One straightforward way is to store the intermediate result in a variable and then use `order()` on that variable: 
```{r}
ans <- ans[order(origin, -dest)]
head(ans)
```

But the previous approach requires having to assign the intermediate result and then overwrite that result. 
We can avoid this intermediate assignment to a temporary variable altergether by *chaining* expressions. 
```{r}
ans <- flights[carrier == "AA", .N, by = .(origin, dest)][order(origin, -dest)]
head(ans, 10)
```

- We can tack expression one from another, forming a chain of operations, i.e., `DT[...][...][...]`. 
- Or we can also chain them vertically: 
```
    DT[...
      ][...
       ][...
        ]
```

### d) Expressions in `by`

If we would like to find out how many flights started late but arrived earlier (or on time), 
started and arrived late, etc...
```{r}
ans <- flights[, .N, by = .(dep_delay > 0, arr_delay > 0)]
ans
```

- The last row corresponds to `dep_delay > 0 = TRUE` and `arr_delay > 0 = FALSE`. 
  We can see that 26593 flights started late but arrived early (on time). 

- Note that we did not provide any names to the `by-expression`. 
  Therefore, names have been automatically assigned in the result. 
  As with `j`, you can name these expressions as you would elements of any `list`, 
  e.g., `DT[, .N, .(dep_delayed = dep_delay > 0, arr_delayed = arr_delay > 0)]`.  

- You can provide other columns along with expressions, for example: 
  `DT[, .N, by = .(a, b > 0)]`. 


### e) Multiple columns in `j`- `.SD` 

#### - Do we have to compute `mean()` for each column individually? 
No necessarily! 
Refresh on th **tip** in Section [1d)](#1d) - *"As long as the `j`- expression returns a `list`,*
*each element of the `list` will be converted to a column in the resulting `data.table`"*. 

Suppose we can refer to the *data subset* for each group as a variable *while grouping*, 
then we can loop through all the columns of that variable using the already- or soon-to-be-familiar 
base function `lapply()`. 

#### Special symbol `.SD`: 
`.SD` stands for **S**ubset of **D**ata. 
It by itself is a `data.table` that holds the data for the *current group* defined by using *by*. 

Let's use the `data.table` `DT` created at the beginning to get a glimpse of what `.SD` looks like. 
```{r}
DT

DT[, print(.SD), by = ID]
```

- `.SD` contains all columns except the *grouping columns* by default.
- It is also generated by preserving the original order - data corresponding to `ID = "b"`, 
  then `ID = "a"`, and then `ID = "c"`. 


To compute on (multiple) columns, we can then simply use the base R function `lapply()`. 
```{r}
DT[, lapply(.SD, mean), by = ID]
```

- `.SD` holds the rows corresponding to columns `a`, `b`, and `c` for that group. 
  We compute the `mean()` on each of these columns using `lapply()`.

- Each group returns a list of three elements containing the mean value which will become the columns 
  of the resulting `data.table`. 

- Since `lapply()` returns a list, so there is no need to wrap it with an additional `.()`. 


#### - How can we specifying just the columns we would like to compute the `mean()` on?

#### .SDcols
Using the argument `.SDcols`. 
It accepts either column names or column indices. 
For example, `.SDcols=c("arr_delay", "dep_delay")` ensures that `.SD` contains only these two columns for each group.  

Similar to part [1g)](#1g), you can also provide the columns to remove instead of columns to keep using `-` or `!` sign 
as well as select consecutive columns as `colA:colB` and deselect consecutive columns as `!(colA:colB)` or `-(colA:colB)`. 

The following code is to use `.SD` along with `.SDcols` to get the `mean()` of `arr_delay` and `dep_delay` columns 
grouped by `origin`, `dest`, and `month`. 

```{r}
flights[carrier == "AA",                        ## Only on trips with carrier "AA"
        lapply(.SD, mean),                      ## compute the mean
        by = .(origin, dest, month),            ## for every 'origin, dest, month'
        .SDcols = c("arr_delay", "dep_delay")]  ## for just those specified in .SDcols
```


### f) Subset `.SD` for each group 

#### - How can we return the first two rows for each `month`? 

```{r}
ans <- flights[, head(.SD, 2), by = month]
head(ans)
```

- `.SD` is a `data.table` that holds all the row for *that group*. 
  We simply subset the first two rows as we have seen somewhere above this page...

- For each group, `head(.SD, 2)` returns the first two rows as a data.table, 
  which is also a `list`, so we do not have to wrap it with `.()`


### g) Why keep `j` so flexible? 

So that we have a consistent syntax and keep using already existing (and familiar) 
base functions instead of learning new functions. 

#### - How can we concatenate columns A and B for each group in `ID`? 
```{r}
DT

DT[, .(val = c(a, b)), by = ID]
```


#### - What if we would like to have all the values of column `a` and `b` concatenated, but returned as a list column? 
```{r}
DT[, .(val = list(c(a, b))), by = ID] 
```

- Here, we first concatenate the values with `c(a, b)` for each group, and wrap that with `list()`. 
  So for each group, we return a list of all concatinated values.

- Note those commas are for display only. 
  A list column can contain any object in each cell, and in this example, 
  each cell is itself a vector and some cells contain longer vectors than others. 

Well, `j` is a powerful syntax..! 
Some example: 
```{r}
## (1) look at the difference between 
DT[, print(c(a, b)), by = ID] 

## (2) and 
DT[, print(list(c(a, b))), by = ID]
```

In (1), for each group, a vector is returned, with length = 6, 4, 2 here. 
However, (2) returns a list of length(1) for each group, with its first element holding vectors of 
length 6, 4, 2. 
Therefore (1) results in a length of `6+4+2=12`, whereas (2) returns `1+1+1=3`.  


---------------------------------------------
## Summary 

The general form of `data.table` syntax is: 
```
DT[i, j, by]
```

#### Using `i`: 

- We can subset rows similarly to a `data.frame`- except we do not have to use `DT$` respectively.   
- We can also sort a `data.table` using `order()` 

We can do much more in `i` by keying a `data.table`, which allows fast subsets and joins. 
We will see this in the 
[*"Keys and fast binary search based subset"*](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html)
and 
[*"Joins and rolling joins"*](https://cran.r-project.org/web/packages/table.express/vignettes/joins.html)  
vignette. 

#### Using `j`:

- Select columns the `data.table` way: `DT[, .(colA, colB)]`.  
  
- Select columns the `data.frame` way: `DT[, c("colA", "colB")]`.  
  
- Compute on the columns: `DT[, .(sum(colA), mean(colB))]`. 

- Provide names if necessary: `DT[, .(sA = sum(colA), mB = mean(colB))]`. 

- Combine with `i`: `DT[colA > value, sum(colB)]`. 


#### Using `by`: 

- Using `by`, we can group by columns by specifying a *list a columns* or *a character vector of column names* or even *expressions*. 

- `by` can handle multiple columns and also *expressions*. 

- `keyby` groups columns to automatically sort the grouped result. 

- Use `.SD` + `.SDcols` in `j` to operate on multiple columns using base functions such as `lapply()` and `head()`. Some examples: 
  - `DT[, lapply(.SD, fun), by = ..., .SDcols = ...]` - applies `fun` to all columns specified in `.SDcols` 
     while grouping by the columns specified in `by`.   
  - `DT[, head(.SD, 2), by = ...]` - return the first two rows for each group. 
  - `DT[col > val, head(.SD, 1), by = ...]` - combine `i` along with `j` and `by`. 

#### Remember the tip: 
As long as `j` returns a `list`, each element of the list will become a column in the resulting `data.table`.

We will see how to *add/update/delete columns by reference* and how to combine them with i and by in the [next vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html).





## Links of vignettes

https://cran.r-project.org/web/packages/data.table/vignettes/

